---
title: "Covid-19 Data Analysis"
author: "Peter Mehler"
date: "Updated April 17th 2020"
output: html_document
---
#Introduction
The dataset used in this analysis is John Hopkins' Covid-19 dataset found here:  https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series. The data is presented in two documents: one for cumulative confirmed cases and the other for cumulative confirmed deaths for most countries worldwide. This information is pulled from a multitude of sources including governments and the WHO, and synthesized in this document. Values are updated daily by the owners of the database.  

Both number of confirmed cases and number of confirmed deaths will be important figures in this analysis. Number of deaths is the more reliable statistic as confirmed cases relies on testing frequency and capacity.  The number of confirmed cases is therefore a lower bound for the number of actual cases of COVID-19.  Changes in confirmed cases could also be due to an increase in testing, so these numbers should be viewed with caution.  

The importance of analyses like these cannot be understated.  After watching the news daily in the last two weeks, this dataset was main source for most of the statistics presented.  The number of confirmed cases, and how fast that number is growing, has been on the forefront of people's minds worldwide.  This dataset wields immense power and is extremely important to determine the severity of the situation that we find ourselves in, and is essential for strategic response.

*Text does not update with data, analysis is as of March 26th*    

```{r, include=FALSE}
confirmed_cases <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv") #,check.names = FALSE to get rid of X prefix

deaths <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv") #,check.names = FALSE to get rid of X prefix

states_data <- read.csv("https://covidtracking.com/api/v1/states/current.csv")
```
```{r, include=FALSE}
#TO DO:

# 1. Animate?
# 1.5 World Map?
# 2. Clean up to make it more report oriented
# 3. Make multiple province countries functional with clean_data()
```

```{r, include=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
```

```{r, include=FALSE}
#state_group <- c("New York", "Washington", "California", "New Jersey", "Massachusetts", "Florida")
```

```{r, include=FALSE}
# us <- confirmed_cases %>% 
#   #filter(Province.State=="New York" | Province.State=="Washington" |Province.State=="California" | #Province.State=="New Jersey" | Province.State=="Massachusetts" | Province.State=="Florida") %>%
#   filter(Country.Region=="US") %>% 
#   gather(key = X1.22.20, value = cases, 5:names(confirmed_cases)[length(names(confirmed_cases))]) %>% 
#   mutate(date = sub("X", "", X1.22.20)) %>% 
#   select(-X1.22.20) %>% 
#   mutate(date = as.Date(date,format="%m.%d.%y")) %>%
#   arrange(-cases)
# 
# # Find locations with top number cases
# temp <- head(subset(us, date==us[["date"]][1]), 10)
# states <- temp$Province.State 
#   
# us <- filter(us, Province.State %in% states)
```

```{r, include=FALSE}
# ggplot(data = us, aes(x=date, y=cases, group = Province.State, color = Province.State, label = Province.State)) +
#   labs(title = "Number of Cases in Top 6 Most Infected US States") +
#   geom_line() +
#   xlim(as.Date("3.8.20",format="%m.%d.%y"),NA) +
#   theme(legend.title = element_blank())
```
```{r, include=FALSE}
# # Plot without New York
# ggplot(data = subset(us, Province.State!="New York"), 
#        aes(x=date, y=cases, group = Province.State, color = Province.State, label = Province.State)) +
#   labs(title = "Number of Cases in Top 6 Most Infected US States") +
#   geom_line() +
#   xlim(as.Date("3.8.20",format="%m.%d.%y"),NA) +
#   theme(legend.title = element_blank())
```

```{r, echo=FALSE}
# Compute power too low to handle master dataset, so individual countries were chosen
# region == 1 means country, region == 2 means province
clean_data <- function(name, region){
   if (region==1){
    temp_deaths <- deaths %>% 
      filter(Country.Region==name) %>% 
      #Convert to long format
      gather(key = X1.22.20, value = deaths, 5:names(confirmed_cases)[length(names(confirmed_cases))]) %>%
      mutate(date = sub("X", "", X1.22.20)) %>%  #get rid of the X prefix
      select(-X1.22.20) %>% #delete old column
      mutate(date = as.Date(date,format="%m.%d.%y")) %>%  #turn it into a date object
      select(deaths, date)
    
    # temp_recovered <- recovered %>% 
    #   filter(Country.Region==name) %>% 
    #   gather(key = X1.22.20, value = recovered, 5:names(confirmed_cases)[length(names(confirmed_cases))]) %>% 
    #   mutate(date = sub("X", "", X1.22.20)) %>% 
    #   select(-X1.22.20) %>% 
    #   mutate(date = as.Date(date,format="%m.%d.%y")) %>% 
    #   select(recovered, date)
      
    temp_cases <- confirmed_cases %>% 
      filter(Country.Region==name) %>% 
      gather(key = X1.22.20, value = cases, 5:names(confirmed_cases)[length(names(confirmed_cases))]) %>% 
      mutate(date = sub("X", "", X1.22.20)) %>% 
      select(-X1.22.20) %>% 
      mutate(date = as.Date(date,format="%m.%d.%y"))
    
    temp_cases <- left_join(temp_cases, temp_deaths, "date")
  #  temp_cases <- left_join(temp_cases, temp_recovered, "date")
    
    temp_cases <- temp_cases %>%
      gather(key = type, value = amount, cases, deaths) # recovered removed
  
        df <- temp_cases %>% 
      arrange(-as.integer(date)) %>%
      arrange(type)
    
    ### calculate slope ###
    for(i in 1:(length(df$amount)-1)){
      if(i!=length(df$amount)/2){ # avoid calculating for change in type from cases to deaths
        df$slope[i] <- df$amount[i]-df$amount[i+1] # might need to mutate slope column
      }
      else{df$slope[i] <- 0}
    }
    df$slope[length(df$amount)] <- 0
    
    # Double derivative
    for(i in 1:(length(df$amount)-1)){
      if(i!=length(df$amount)/2){ # avoid calculating for change in type from cases to deaths
        df$double_derivative[i] <- df$slope[i]-df$slope[i+1] # might need to mutate slope column
      }
      else{df$double_derivative[i] <- 0}
    }
    df$double_derivative[length(df$amount)] <- 0
    
    return(df)
  }
  
  else{
    temp_deaths <- deaths %>% 
      filter(Province.State==name) %>% 
      gather(key = X1.22.20, value = deaths, 5:names(confirmed_cases)[length(names(confirmed_cases))]) %>% 
      mutate(date = sub("X", "", X1.22.20)) %>% 
      select(-X1.22.20) %>% 
      mutate(date = as.Date(date,format="%m.%d.%y")) %>% 
      select(deaths, date)
    
    # temp_recovered <- recovered %>% 
    #   filter(Province.State==name) %>% 
    #   gather(key = X1.22.20, value = recovered, 5:names(confirmed_cases)[length(names(confirmed_cases))]) %>% 
    #   mutate(date = sub("X", "", X1.22.20)) %>% 
    #   select(-X1.22.20) %>% 
    #   mutate(date = as.Date(date,format="%m.%d.%y")) %>% 
    #   select(recovered, date)
      
    temp_cases <- confirmed_cases %>% 
      filter(Province.State==name) %>% 
      gather(key = X1.22.20, value = cases, 5:names(confirmed_cases)[length(names(confirmed_cases))]) %>% 
      mutate(date = sub("X", "", X1.22.20)) %>% 
      select(-X1.22.20) %>% 
      mutate(date = as.Date(date,format="%m.%d.%y"))
    
    temp_cases <- left_join(temp_cases, temp_deaths, "date")
#   temp_cases <- left_join(temp_cases, temp_recovered, "date")
    
    temp_cases <- temp_cases %>%
      gather(key = type, value = amount, cases, deaths) # removed recovered
    
    df <- temp_cases %>% 
      arrange(-as.integer(date)) %>%
      arrange(type)
    
    ### calculate slope ###
    for(i in 1:(length(df$amount)-1)){
      if(i!=length(df$amount)/2){ # avoid calculating for change in type from cases to deaths
        df$slope[i] <- df$amount[i]-df$amount[i+1] # might need to mutate slope column
      }
      else{df$slope[i] <- 0}
    }
    df$slope[length(df$amount)] <- 0
    
    # Double Derivative
    for(i in 1:(length(df$amount)-1)){
      if(i!=length(df$amount)/2){ # avoid calculating for change in type from cases to deaths
        df$double_derivative[i] <- df$slope[i]-df$slope[i+1] # might need to mutate slope column
      }
      else{df$double_derivative[i] <- 0}
    }
    df$double_derivative[length(df$amount)] <- 0
    
    return(df)
  }
}
```


```{r, echo=FALSE}
# # CHOOSE YOUR COUNTRY OR PROVINCE
# 1. Only works for Provinces and Countries without several provinces
# 2. region == 1 means country, region == 2 means province
# data sets
hubei <- clean_data("Hubei", 2)
italy <- clean_data("Italy", 1)
s.korea <- clean_data("Korea, South", 1)
#ny <- clean_data("New York", 2)
iran <- clean_data("Iran", 1)
spain <- clean_data("Spain",1)
us <- clean_data("US",1)
germany <- clean_data("Germany",1)
```
```{r, echo=FALSE}
# Row combine all datasets
combined.countries <- rbind(hubei, italy, s.korea, iran, spain, us)
```


```{r, include=FALSE}
# Works for countries with several provinces
# us_total <- confirmed_cases %>% 
#   filter(Country.Region=="US") %>% 
#   gather(key = X1.22.20, value = cases, 5:names(confirmed_cases)[length(names(confirmed_cases))]) %>% 
#   mutate(date = sub("X", "", X1.22.20)) %>% 
#   select(-X1.22.20) %>% 
#   mutate(date = as.Date(date,format="%m.%d.%y")) %>%
#   gather(key = type, value = amount, cases) %>% 
#   group_by(date) %>% 
#   summarize(amount = sum(amount),
#             Country.Region = "US")
```

```{r, include=FALSE}
hubei %>% 
  filter(type=="cases") %>% 
  ggplot(aes(x=date, y=amount, group = 1)) +
  labs(title = "hubei province, red line indicates Wuhan shutdown") +
  geom_line() + 
  geom_vline(xintercept = as.Date("1.23.20",format="%m.%d.%y"), 
                color = "red", size=1) +
  xlim(as.Date("1.22.20",format="%m.%d.%y"),NA)
  
#Wuhan Lockdown January 23rd
```
#Results

Let's begin by looking at the progression of Hubei province.  With strict contact tracing, complete lockdown of all non-essential business, and a travel ban, China was able to stem the flow of the rate of infection in about eight weeks. The death rate hovers just below 4% for China. This pattern follows the standard S curve of infectious diseases where infection rate begins exponential, then tapers off.  
```{r, echo=FALSE}
ggplot(data = hubei, aes(date, amount, fill = type)) +
  geom_bar(position="stack", stat = "identity") +
  labs(title = "Cases, Deaths in Hubei Province Through Time",
       x = "Date",
       y = "Amount") +
  theme(legend.title=element_blank())

```
  
Let's look at Italy.One look at the number of cases on the left side and it becomes clear that this graph is worse than the previous.  Italy's new daily cases are seeing a drop, we take a look at the rate of the drop in the next graph.    
```{r, echo=FALSE}
ggplot(data = italy, aes(date, amount, fill = type)) +
  geom_bar(position="stack", stat = "identity") +
  theme(legend.title=element_blank()) +
  labs(title = "Cases and Deaths in Italy Through Time",
       x = "Date",
       y = "Amount")
```
  
Here, a number above 0 suggests an exponential increase in deaths and new cases, and a number below 0 suggests a falling rate of infection. The rate at which the number of new reports are occuring is falling.  Below 0, the disease can, in one respect, be considered under control, because the number of new cases is not building upon itself exponentially.   
```{r, echo=FALSE, message=FALSE}
italy %>% 
  #filter(type=="cases") %>% 
  ggplot(aes(x=date, y=double_derivative, group = type, color = type)) +
  labs(title = "Rate of Rate of Increase of Deaths and Cases in Italy", 
       caption = "Error is hidden, but raw data was fairly volatile and the standard deviation is large.  Pattern remains meaninful and more rigorous analysis is necessary",
       x = "Date",
       y = "Daily Increase of Reports Compared to Previous Day") +
  #geom_line() + 
  geom_smooth(se=F, na.rm = T) +
  theme(legend.title=element_blank()) +
   xlim(as.Date("2.15.20",format="%m.%d.%y"),NA)
```
  
```{r, include=FALSE}
us %>% 
  filter(type=="cases") %>% 
  ggplot(aes(x=date, y=double_derivative, group = 1)) +
  labs(title = "US", caption = "caption" ) +
  #geom_line() + 
  geom_smooth() +
  theme_minimal() +
   xlim(as.Date("2.15.20",format="%m.%d.%y"),NA)
```
Let's look at the same plot, but for other countries and without deaths.   
```{r, echo=FALSE, message=FALSE}
combined.countries %>% 
  filter(type=="cases") %>% 
  ggplot(aes(x=date, y=double_derivative, group = Country.Region, color = Country.Region)) +
  labs(title = "Rate of Rate of Increase of Confirmed Cases by Country", 
       caption = "Standard deviation is significant  Pattern remains meaninful but more rigorous analysis is necessary",
       x = "Date",
       y = "Daily Increase of Reports Compared to Previous Day") +
  #geom_line() + 
  geom_smooth(se=F) +
  theme_minimal()
  #xlim(as.Date("1.23.20",format="%m.%d.%y"),NA)
```
  
Here there are two major trends I would like to point out.  The first is obviously how dramatic the rate of increase in the United States was.  Caught unprepared, the Virus was able to establish a foothold in NYC, a city with high population density, before society and the government implimented proper spread suppression procedures.  

The other example I wish to highlight is Hubei's (Labeled China in the plot) line. The completed curve of the rate of the rate of increase will meander through the x-axis, past 0, and into negative numbers until new cases dwindle to almost 0 (barring another outbreak, which is still possible).  

Here is the same plot, but for deaths.  

```{r, echo=FALSE, message=FALSE}
combined.countries %>% 
  filter(type=="deaths") %>% 
  ggplot(aes(x=date, y=double_derivative, group = Country.Region, color = Country.Region)) +
  labs(title = "Rate of Rate of Increase of Deaths by Country", 
       caption = "Standard deviation is significant.  Pattern remains meaninful but more rigorous analysis is necessary",
       x = "Date",
       y = "Daily Increase of Reports Compared to Previous Day") +
  #geom_line() + 
  geom_smooth(se=F) +
  theme_minimal()
  #xlim(as.Date("1.23.20",format="%m.%d.%y"),NA)
```
  
 
```{r, include=FALSE}
# Let's look at death rates as percentages of cases.  Here I compare a country with one of the highest case fatality rates (Italy), with a country with one of the lowest (Germany).  It should be noted that this is an unfair comparison at the current date of writing as Italy has been dealing with the virus for a few weeks longer than Germany. 
require(gridExtra)
```

```{r, include=FALSE}
plot1 <- ggplot(data = italy, aes(date, amount, fill = type)) +
  geom_bar(position="fill", stat = "identity") +
  labs(title = "Percentages of Case Deaths Italy", y="Percent", x="Date") +
  theme(legend.title=element_blank())
plot2 <- ggplot(data = germany, aes(date, amount, fill = type)) +
  geom_bar(position="fill", stat = "identity") +
  labs(title = "Percentages of Case Deaths Germany", y="Percent", x="Date") +
  theme(legend.title=element_blank())

grid.arrange(plot1, plot2, ncol=2)
```
  
Below are the plots for cumulative deaths and cases since the beginning of the outbreak in various countries.  
```{r, echo=FALSE, message=FALSE}
combined.countries %>%
  filter(type=="cases") %>%   # Have filtered by cases
  filter(amount>400) %>%    # Filter cases > 400
  group_by(Country.Region) %>% 
  mutate(date = as.integer(format(date, "%j"))) %>% 
  mutate(date = date - date[length(date)]) %>% # Change date to days since cases > 400
  ggplot(aes(date, amount, color = Country.Region)) +
    # geom_line() +
    # geom_vline(xintercept = 1, color = "red", size=.25) + # Wuhan locks down
    # geom_vline(xintercept = 11, color = "green", size=.25) + # Italy locks down
    # geom_vline(xintercept = -1, color = "blue", size=.25) + # South Korea Socially Isolates
    # geom_vline(xintercept = 8, color = "purple", size=.25) + # NY State orders lockdown
    labs(x="Days since number of cases reached 400", 
         y="Cumulative Cases", 
         title = "Comparison of Cumulative Cases by Country",
         subtitle = "Beginning When Number of Infections is Above 400",
         caption = "China represents only Hubei data") +
    geom_smooth(size=.75, na.rm = T) +
    ylim(0,NA) +
    theme(legend.title=element_blank())
                        

#Majority of people of South Korea decide to socially isolate 2.20.20 
#Italy Lockdown 3.8.20
#Wuhan Lockdown 1.23.20
#NY State orders lockdown
```
   

```{r, include=FALSE}
# How to determine Slope?  Doubling every ___ days?
combined.countries %>% 
  filter(type=="cases") %>% 
  filter(amount>400) %>%    # Filter cases > 400
  group_by(Country.Region) %>% 
  mutate(date = as.integer(format(date, "%j"))) %>% 
  mutate(date = date - date[length(date)]) %>% # Change date to days since first death
  ggplot(aes(date, amount, color = Country.Region)) +
    geom_line() +
    labs(x="Days since number of cases reached 400", 
         y="Cumulative Cases", 
         title = "Exponential Projections of Cumulative Cases by Country",
         subtitle = "Beginning When Number of Infections is Above 400",
         caption = "China represents only Hubei data") +
    #xlim(NA,as.Date(50, origin="2019-12-31")) +
    ylim(0,NA) +
    #geom_smooth(size=.75) +
    theme(legend.title=element_blank()) +
    scale_y_log10() +
    geom_abline(slope=.1, intercept=3)
    #stat_smooth(method="lm",fullrange=TRUE) 
```

```{r, echo=FALSE, message=FALSE}
# Row combine all datasets
combined.countries %>%
  filter(type=="deaths") %>%   # Have filtered by deaths
  filter(amount>0) %>%    # Filter cases > 0
  group_by(Country.Region) %>% 
  mutate(date = as.integer(format(date, "%j"))) %>% 
  mutate(date = date - date[length(date)]) %>% # Change date to days since first death
  ggplot(aes(date, amount, color = Country.Region)) +
    #geom_line() +
    labs(x="Days since first Death", 
         y="Cumulative Deaths", 
         title = "Comparison of Cumulative Deaths by Country Beginning After First Death", 
         caption = "China represents only Hubei data") +
    ylim(0,NA) +
    geom_smooth(size=.75, na.rm = T) +
    theme(legend.title=element_blank())
```
  
```{r, include=FALSE}
combined.countries %>%
  filter(type=="deaths") %>%   # Have filtered by deaths
  filter(amount>0) %>%    # Filter cases > 0
  group_by(Country.Region) %>% 
  mutate(date = as.integer(format(date, "%j"))) %>% 
  mutate(date = date - date[length(date)]) %>% # Change date to days since first death
  ggplot(aes(date, amount, color = Country.Region)) +
    #geom_line() +
    labs(x="Days since first Death", 
         y="Cumulative Deaths", 
         title = "Exponential Projections of Cumulative Deaths by Country Beginning After First Death", 
         caption = "China represents only Hubei data") +
    ylim(0,NA) +
    #geom_smooth() +
    theme(legend.title=element_blank()) +
    stat_smooth(method="lm",fullrange=TRUE)
```

```{r, include=FALSE}
# add summed weekly column to prep for graph below
combined.countries.cumulative <- combined.countries %>% 
  filter(type=="cases") %>% 
  group_by(Country.Region) %>% 
  arrange( Country.Region, -as.integer(date))
```
```{r, include=FALSE}
# adds weekly column which is a weekly sum of the slope column "Number of New Daily Cases in the Past Week"
num.days = length(names(confirmed_cases))-4

for(i in 1:(length(combined.countries.cumulative$slope))){
  if(num.days-(i%%num.days) < 7){
    combined.countries.cumulative$weekly[i] = sum(combined.countries.cumulative$slope[i:(i+num.days-(i%%num.days))])
  }
  else if(i%%num.days == 0){
    combined.countries.cumulative$weekly[i] =combined.countries.cumulative$slope[i]
  }
  else{
    combined.countries.cumulative$weekly[i] = sum(combined.countries.cumulative$slope[i:(i+6)])
  }
}
```
Which countries are experiencing exponential growth?  The following graph, inspired by Aatish Bhatia (https://aatishb.com/covidtrends/), graphs cumulative cases against daily new cases summed over the week prior to that day. The straighter the line of a given country, the more similar to true exponential growth in deaths or cases that country has.  The countries you see falling off that straight line pattern are beginning to see a slow down in the number of new cases.
```{r, echo=FALSE}
combined.countries.cumulative %>%
  filter(amount>400) %>%    # Filter cases > 400
  group_by(Country.Region) %>% 
  mutate(date = as.integer(format(date, "%j"))) %>% 
  mutate(date = date - date[length(date)]) %>% # Change date to days since cases > 400
  ggplot(aes(amount, weekly, color = Country.Region)) +
    geom_line() +
    labs(x="Cumulative Cases", 
         y="Number of New Daily Cases in the Past Week", 
         title = "Which countries have exponential case growth?",
         subtitle = "Beginning When Number of Infections is Above 400",
         caption = "China represents only Hubei data") +
    #geom_smooth(size=.5, se=F) +
    scale_y_log10() +
    scale_x_log10() +
    theme(legend.title=element_blank())

    
```
  
```{r, include=F}
library(maps)
library(openintro)
#world_map <- map_data("world")  #to map countries in the world
states_map <- map_data("state") #to map states in the US
```
```{r, include=F}
states_data <- states_data %>% 
  mutate(state = tolower(abbr2state(state))) %>% 
  rename(region = state)
```
```{r, include=F}
states_geo <- left_join(states_map, states_data, by = "region")
```
Maps below are unfinished, and I am still playing around with them!
```{r, echo=F}
states_geo %>%
  ggplot(aes(long, lat, fill=positive)) +
    labs(title = "Confirmed Cases of COVID-19", x = "Longitude", y = "Latitude") +
    geom_polygon(aes(group=group), col = "black") +
    coord_fixed(ratio = 1.15)  +
    scale_fill_gradientn(colors = c('blue','yellow', 'red')) + #customize colors  colors = brewer.pal(13, "Spectral")
    guides(fill=guide_legend(title="Confirmed Cases"))
```  
```{r, echo=F}
states_geo %>%
  ggplot(aes(long, lat, fill=death)) +
    labs(title = "Deaths from COVID-19", x = "Longitude", y = "Latitude") +
    geom_polygon(aes(group=group), col = "black") +
    coord_fixed(ratio = 1.15)  +
    scale_fill_gradientn(colors = c('blue','yellow', 'red')) + #customize colors  colors = brewer.pal(13, "Spectral")
    guides(fill=guide_legend(title="Deaths"))
```
  
```{r, include=F}
library(tidycensus)
library(RColorBrewer)
```
  
```{r, include=F}
nyc_counties <- c("Kings County, New York","New York County, New York", "Queens County, New York", "Richmond County, New York", "Bronx County, New York")

nyc_geo <- get_acs(geography = "county",
                        state = "New York", # if we only want NY counties
                        table = "B03001",
                        year = 2018,
                        survey = "acs5",
                        geometry = TRUE)


```
```{r, include=F}
boroughs <- read.csv("https://raw.githubusercontent.com/nychealth/coronavirus-data/master/boro.csv")
```

```{r, include=F}
boroughs <- boroughs %>% 
  filter(BOROUGH_GROUP != "Citywide") %>% 
  arrange(BOROUGH_GROUP)

boroughs$NAME <- nyc_counties
```

```{r, include=F}
nyc_geo <- nyc_geo %>% 
  filter(NAME %in% nyc_counties)
```

```{r, include=F}
nyc_geo <- left_join(nyc_geo, boroughs, by = "NAME")
```
```{r, echo=F}
nyc_geo %>% 
  ggplot(aes(fill=COVID_CASE_RATE)) +
  geom_sf() +
  geom_sf_text(aes(label = BOROUGH_GROUP), fontface="bold") +
  labs(title = "Rate of COVID-19 Cases by Borough", subtitle = "Rate = Cases per 100,000 residents", x = "Latitude", y = "Longitude") +
  scale_fill_gradientn(colors = brewer.pal(9, "YlOrRd")) + #customize colors  colors = brewer.pal(13, "Spectral")
  guides(fill=guide_legend(title="Rate"))
```
```{r, echo=F}
nyc_geo %>% 
  ggplot(aes(fill=COVID_CASE_COUNT)) +
  geom_sf() +
  geom_sf_text(aes(label = BOROUGH_GROUP), fontface="bold") +
  labs(title = "COVID-19 Cases by Borough", x = "Latitude", y = "Longitude") +
  scale_fill_gradientn(colors = brewer.pal(9, "YlOrRd")) + #customize colors  colors = brewer.pal(13, "Spectral")
  guides(fill=guide_legend(title="Count"))
```
  

#Discussion
The conclusions of this report are changing daily with the updates of the dataset. Centralized datasets such as these are not the only information source which should inspire action, but they can play a significant role.  Without data analysis during times of crisis such as this, we are relying on the individual beliefs of people. When an issue's scope is at the level of the entire world, no amount of intuition will ever surpass the insights of large data.  

This is not to say that data is infalliable.  When looking into the sources of data and methods of data collection for this dataset, It becomes clear that it is far from perfect.  Each country's numbers rely on the government's honor that reports are valid, and assumptions must be made about the confidence of data collection at multiple levels of the pipeline.  

Further work is being relentlessly pursued globally. As more data come out, more questions are able to be answered.  We are still unsure of the exact case fatality rate, the exact spread rate, or, perhaps most importantly, what measures best defend human life.

Aknowledgements:  

https://aatishb.com/covidtrends/

https://medium.com/@tomaspueyo/coronavirus-the-hammer-and-the-dance-be9337092b56?fbclid=IwAR2xNhWfHPb1189m7moSCqS52LdWESHPF0UfXN8lf8ogZetYWUD8EONyGqY  

https://medium.com/@tomaspueyo/coronavirus-act-today-or-people-will-die-f4d3d9cd99ca

https://stackoverflow.com/questions/51355303/extract-slope-of-multiple-trend-lines-from-geom-smooth    

https://www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-NPI-modelling-16-03-2020.pdf  

https://www.youtube.com/watch?v=Kas0tIxDvrg  

https://www.youtube.com/watch?v=54XLXg4fYsc&feature=share&fbclid=IwAR2MKASftMjgiAZgXzCF05iccj1nn2tA3H13LCP1EvDHyhXv1LUgPblVD-U